---
title: "1/_f_ noise is related to age, but not short-term memory or sensorimotor reaction time."
author: "Jan Freyberg"
email: jan.freyberg@kcl.ac.uk
department: Forensic and Neurodevelopmental Sciences, King's College London
address: 16 De Crespigny Park, SE5 8AF London, UK
date: "12/05/2017"
output:
  bookdown::pdf_document2:
    template: ccn_style.tex
    fig_caption: true

params:
  nboot: 5000
urlcolor: blue
header-includes:
  - \usepackage{hyperref}
  - \usepackage{float}

abstract: 
  "Recent evidence points towards an increase in 1/_f_ neural noise (noise with decreasing power in higher frequency bands) with age. Previous evidence related this increase to a decrease in working memory performance. Here, we attempted to replicate previous findings with data from the Cam-CAN project. We used two minutes of a resting-state recording to estimate the power-spectral density of neural activity. We then correlated the $\\mathbf{\\beta}$ of a regression of noise on frequency with age, sensorimotor reaction times, and short-term memory performance. We reproduced previously demonstrated increases in 1/_f_ noise with age, but although 1/_f_ noise mediated a relationship between age and working memory, it did not independently predict working memory performance. These results indicate that 1/_f_ noise is a result of ageing, but that may contribute little to cognitive effects of ageing."

keywords: MEG; noise; ageing
bibliography: /Users/jan/Documents/Zotero/better-bibtex/zotero-library.bib
---

```{r setup, include=FALSE}
# for plotting:
library(png)
library(grid)
library(gridExtra)
library(jpeg)
library(ggplot2)
library(viridis)

# for printing
library(apa)
library(stringr)

# for data processing:
library(readr)
library(stringr)
library(dplyr)
library(tidyr)
library(magrittr)
library(mediation)

# load the data for the two MEG recordings
task.data <- read_csv('../csv/task.csv') %>%
  filter(removed_elecs < 1.2*204)
rest.data <- read_csv('../csv/rest.csv') %>%
  filter(removed_elecs < 1.2*204)

# add the MEG behavioral data
task.data <- inner_join(task.data, read_csv('../csv/megbehavior.csv'))
# add the VSTM behavioral data
task.data <- read_csv('../csv/vstm-behavior.csv') %>%
  rename(pid = CCID) %>%
  mutate_each(funs(as.numeric), -pid) %>%
  inner_join(task.data)

# load the 
both.data <- inner_join(task.data, rest.data, by=c('pid', 'sex', 'age'), suffix=c('.task', '.rest'))

# load the spectrum data for age split:
age.split.spectra <- read_csv('../csv/agesplit-spectra.csv') %>%
  gather(frequency, power, -pid, -group) %>%
  mutate(frequency = as.numeric(frequency)) %>%
  mutate(group = str_replace(group, 'old', 'Oldest 25%')) %>%
  mutate(group = str_replace(group, 'young', 'Youngest 25%')) %>%
  mutate(group = factor(group, levels = c('Youngest 25%', 'Oldest 25%')))

# helper function
fmt_tex <- function(text){
  str_replace(paste0("$", text, "$"), ', ', '$, $') %>%
    str_replace('%', '\\\\%')
}

age.slope.task <- task.data %$% cor.test(age, average.slope, method='p') %>% apa(text, format='text') %>% fmt_tex()
age.slope.rest <- rest.data %$% cor.test(age, average.slope, method='p') %>% apa(text, format='text') %>% fmt_tex()

age.intercept.cor <- task.data %$% cor.test(age, average.intercept, method='p') %>% apa(text, format='text') %>% fmt_tex()
rt.slope.cor <- task.data %$% cor.test(mean.rt, average.slope, method='p') %>% apa(text, format='text') %>% fmt_tex()
vstm.slope.cor <- task.data %$% cor.test(Prcsn_ss2, average.slope, method='p') %>% apa(text, format='text') %>% fmt_tex()

rest.slope.task.cor <- both.data %$% cor.test(average.slope.rest, average.slope.task, method='p') %>% apa(text, format='text') %>% fmt_tex()

### Run a mediation analysis
mediationresults <- mediate(lm(average.slope.rest ~ age, data=both.data),
                            lm(formula = Prcsn_ss2 ~ average.slope.rest + age, data = both.data),
                            treat='age', mediator='average.slope.rest', boot=TRUE, sims=params$nboot)
total.eff <- paste0('b=', signif(mediationresults$tau.coef, digits=3), ', ',
                    # ', 95% CI: ', signif(mediationresults$tau.ci[1], digits=3),
                    # '-', signif(mediationresults$tau.ci[2], digits=3), ', ',
                    ifelse(mediationresults$tau.p<0.001, 'p<0.001', paste0('p=', round(mediationresults$tau.p, digits=3)))) %>%
  fmt_tex()
direct.eff <- paste0('b=', signif(mediationresults$z0, digits=3), ', ',
                    # '95% CI: ', signif(mediationresults$z0.ci[1], digits=3),
                    # '-', signif(mediationresults$z0.ci[2], digits=3), ', ',
                    ifelse(mediationresults$z0.p<0.001, 'p<0.001', paste0('p=', round(mediationresults$z0.p, digits=3)))) %>%
  fmt_tex()
indirect.eff <- paste0('b=', signif(mediationresults$d0, digits=3), ', ',
                    # '95% CI: ', signif(mediationresults$d0.ci[1], digits=3),
                    # '-', signif(mediationresults$d0.ci[2], digits=3), ', ',
                    ifelse(mediationresults$d0.p<0.001, 'p<0.001', paste0('p=', round(mediationresults$d0.p, digits=3)))) %>%
  fmt_tex()
```

# Introduction

One feature of the power spectrum of human brain activity obtained through electrophysiological imaging is a decrease in power with increasing frequency. This relationship between power and the inverse of frequency (often described as 1/_f_, or pink, noise) has recently been linked to ageing [@voytek_age-related_2015]. Voytek and colleagues found that the linear relationship between frequency and noise was steeper in younger individuals, with less electrophysiological noise in high frequency bands. They were able to show this relationship using both intracranial electrocorticography (ECoG) and electroencephalography (EEG). Voytek and colleagues also demonstrated a relationship between the levels of 1/_f_ noise and visual working memory, and that 1/_f_ noise mediates impairments of visual working memory in ageing.

Here, we wanted to reproduce the finding and test its central hypoethis of a relationship between  in a public dataset: the Cambridge Center for Ageing and Neuroscience data repository [Cam-CAN, @taylor_cambridge_2017]. We therefore tried to replicate the finding of 1/_f_ noise increasing in age, as well as the relationship between 1/_f_ noise and working memory performance.

# Methods

The Cam-CAN data includes one 8m40s resting-state and one 8m40s sensorimotor task MEG recordings for 647 participants. Resting-state recordings were conducted with eyes closed, and the sensorimotor task required button response to bilateral sine grating and concurrent audio tone presentation. Participants also completed behavioural tasks outside the MEG scanner. Here, we analysed a visual short term memory (VSTM) task: matching a sequence of hues, storing 1 to 4 hues in short-term memory.

MEG recording methods are fully described in Taylor et al. [-@taylor_cambridge_2017]. Out of 647, we analysed `r nrow(rest.data)` resting-state and `r nrow(task.data)` task recordings. This subset of subjects had an average age of `r task.data %$% mean(age) %>% round(digits=1)` years (st.dev.: `r task.data %$% sd(age) %>% round(digits=1)`, range: `r task.data %$% min(age)`-`r task.data %$% max(age)`).

All data analysis was performed using `MNE-python` [@gramfort_mne_2014]. We chose a random two-minute segment from the full recording. We performed temporal space-signal separation [tSSS, @taulu_applications_2005, correlation threshold 0.98; 10s sliding window], followed by ICA decomposition and removal of ICA components based on the highest correlation with the HEOG, VEOG and ECG channels. We then computed the log of the power-spectrum from 2 Hz to 7 Hz and from 14 Hz to 24 Hz, avoiding the alpha range, using Welch's method (2s sliding window, 50% overlap). We calculated a linear regression of this semi-log data for each gradiometer for each individual and removed any sensors for which the intercept of this linear regression was two standard-deviations above or below the average for an individual was removed.

# Results

The relationship between 1/_f_ noise slope and age replicated both during rest (`r age.slope.rest`) and task (Figure \@ref(fig:taskagecorr), `r age.slope.task`). The 1/_f_ slope during the two recordings predicted each other, although 1/_f_ noise fell more steeply during task (Figure \@ref(fig:resttaskcorr), `r rest.slope.task.cor`). We also found a strong relationship between the 1/_f_ noise slope during resting state and task (Figure \@ref(fig:resttaskcorr), `r rest.slope.task.cor`).

There was no evidence for a relationship between 1/_f_ noise and VSTM (`r vstm.slope.cor`) or sensorimotor reaction time (`r rt.slope.cor`). However, in a mediation analysis with age as predictor and 1/_f_ slope as mediator, we found that 1/_f_ noise partially mediated a strong relationship between age and VSTM (total effect: `r total.eff`; indirect effect: `r indirect.eff`)

We also analysed the relationship between 1/_f_ noise and age at each sensor for each recording (Figure \@ref(fig:topoplot)). We found differences between the two recordings in two clusters (Figure \@ref(fig:diffplot)).

```{r taskagecorr, echo=FALSE, fig.width=3.5, fig.height=2, fig.pos='H', fig.align='left', fig.cap="\\textit{Relationship between 1/f noise during a sensorimotor task and age.}"}
p <- ggplot(task.data, aes(x=age, y=average.slope, color=age)) +
  geom_smooth(method=lm, color='gray') +
  labs(y = "Average slope (task)", color='Age\n', x = "Age (years)") +
  geom_point(size=1.2, alpha=0.6, show.legend = FALSE) +
  scale_color_viridis() +
  theme_bw() +
  scale_x_continuous(expand = c(0, 0)) +
  theme(axis.text.y = element_text(face="bold", size=6, angle=90),
        axis.text.x = element_text(face="bold", size=6),
        legend.position = 'bottom')

p
```

```{r resttaskcorr, echo=FALSE, fig.width=3.5, fig.height=2, fig.pos='H', fig.align='left', fig.cap="\\textit{Relationship between 1/f slope during task and rest.}"}
limits <- both.data %$% c(min(min(average.slope.task), min(average.slope.rest)),
                          max(max(average.slope.task), max(average.slope.rest)))
lmdat <- tibble(x=limits, y=limits)

p <- ggplot(both.data, aes(x=average.slope.rest, y=average.slope.task, color=age)) +
  geom_smooth(method=lm, color='gray') +
  labs(x = "Average slope (task)", y = "Average slope (rest)", color='Age\n') +
  geom_point(size = 1.2, alpha = 0.6, show.legend = FALSE) +
  scale_color_viridis() +
  theme_bw() +
  geom_line(data = lmdat, aes(x = x, y = y), inherit.aes = FALSE) +
  coord_cartesian(xlim = limits, ylim = limits) +
  scale_x_continuous(expand = c(0, 0)) +
  theme(axis.text.y = element_text(face="bold", size=6, angle=90),
        axis.text.x = element_text(face="bold", size=6),
        legend.position = 'bottom')

p
```

```{r topoplot, fig.align='center', out.width='49%', fig.pos='H', echo=FALSE, fig.show='hold', fig.cap="\\textit{Correlation coefficients between age and 1/f noise for each sensor. Sensors for which this relationship was sigificant at cluster-corrected alpha level 0.05 are indicated by white circles.}"}
knitr::include_graphics(c("figures/r-rest.pdf", "figures/r-task.pdf", "figures/r-cbar.pdf"))
```

```{r diffplot, fig.align='center', out.width=c('50%', '40%'), out.extra=sprintf('angle=%s', c(0, 90)), fig.pos='H', echo=FALSE, fig.show='hold', fig.cap="\\textit{The difference between the correlation of 1/f noise and age at each sensor. Sensors for which bootstrapped estimates of this difference were significant are highlighted.}"}
knitr::include_graphics(c("figures/r-difference.pdf", "figures/r-difference-cbar.pdf"))
```

# Conclusions

This analysis replicates reports of increased 1/_f_ noise with age. We were unable to replicate a correlation between 1/_f_ noise with VSTM, and also found no relationship between 1/_f_ noise and sensorimotor reaction times. The difference between this and previous reports may stem from the fact that the MEG recordings took place separately to the VSTM task. Physiological noise occurring during the memory task would likely affect performance on that task more strongly. We believe that this analysis significantly adds to our understanding of how 1/_f_ noise in old age affects cognitive performance.

# References

```{r, echo=FALSE, eval=FALSE, fig.width=3.3, fig.height=2.5, warning=FALSE, fig.cap="\textit{No relationship between 1/f noise and reaction times.}"}
p <- ggplot(task.data, aes(x=mean.rt, y=average.slope, color=age)) +
  geom_smooth(method=lm, color='gray') +
  geom_point(size=1.2, show.legend=FALSE) +
  scale_color_viridis() +
  labs(x = "Sensorimotor Reaction Time (s)", y = "Average Slope", color='Age\n') +
  theme_bw() +
  scale_x_continuous(expand = c(0, 0)) +
  theme(axis.text.y = element_text(face="bold", size=6, angle=90),
        axis.text.x = element_text(face="bold", size=6),
        legend.position = 'bottom')

p
```

```{r, echo=FALSE, eval=FALSE, fig.width=3.3, fig.height=2.5, warning=FALSE, fig.cap="No relationship between 1/f noise and short-term memory."}
p <- task.data %>%
  filter(Prcsn_ss4 <= 1) %>%
  ggplot(aes(x=Prcsn_ss4, y=average.slope, color=age)) +
  geom_smooth(method=lm, color='gray') +
  geom_point(size=1.2, show.legend=FALSE) +
  scale_color_viridis() +
  labs(x = "Sensorimotor Reaction Time (s)", y = "Average Slope", color='Age\n') +
  theme_bw() +
  scale_x_continuous(expand = c(0, 0)) +
  theme(axis.text.y = element_text(face="bold", size=6, angle=90),
        axis.text.x = element_text(face="bold", size=6),
        legend.position = 'bottom')

p
```

```{r, echo=FALSE, eval=FALSE, fig.width=3.3, fig.height=2.6, fig.cap="Power spectra in the top and bottom age quartiles."}
p <- ggplot(age.split.spectra, aes(x=frequency, y=power, group=group, color=group)) +
  stat_summary(fun.y='mean', geom = 'line') +
  geom_ribbon(stat='summary', fun.data='mean_se',
              aes(fill=group), alpha=0.3, size=0,
              show.legend = FALSE) +
  labs(group = '', x = 'Frequency', y = 'Power') +
  theme_bw() +
  scale_fill_manual(values = viridis(2, begin=0.25, end=0.75)) +
  scale_color_manual(values = viridis(2, begin=0.125, end=0.875)) +
  scale_x_continuous(expand = c(0, 0)) +
  theme(axis.text.y = element_text(face="bold", size=6, angle=90),
        axis.text.x = element_text(face="bold", size=6),
        legend.title = element_blank(),
        legend.position = c(0.7, 0.85),
        legend.background = element_blank(),
        legend.box.background = element_blank())

p
```